---
title: "Fitting and comparing detection functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting and comparing detection functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_mrds <- requireNamespace("mrds", quietly = TRUE)
```

```{r, echo = FALSE, results = "asis", eval = !has_mrds}
cat("> **`mrds` is not installed**, so the fitted output below is not shown.\n",
    "> Install it with `install.packages(\"mrds\")` to run this through.\n",
    sep = "")
```

This walks one survey from a table of distances to a `g(0)` correction, stopping
where the package stops. The through-line is that **a detection function is
chosen for its consequences, not its AIC**, and that most of the ways this goes
wrong are silent.

```{r}
library(dsfit)
```

## A survey to work with

No detection data ships with the package, so here is a simulated aerial survey.
It has a blind spot beneath the aircraft, which is what makes it worth using:
that single feature drives most of the decisions below.

```{r}
set.seed(20260810)

# True perpendicular distances of animals present, out to 3 km
present <- abs(rnorm(4000, 0, 900))
present <- present[present < 3000]

# Sea state, which makes animals harder to see
beaufort <- sample(0:4, length(present), replace = TRUE)

# Detection falls off with distance, falls off further in a rough sea, and is
# zero in the 100 m the aircraft cannot see beneath itself
sigma <- 800 - 60 * beaufort
p <- exp(-present^2 / (2 * sigma^2))
p[present < 100] <- 0

seen <- runif(length(present)) < p

detections <- data.frame(
  object   = seq_len(sum(seen)),
  distance = present[seen],
  beaufort = beaufort[seen],
  size     = sample(1:3, sum(seen), replace = TRUE, prob = c(.6, .3, .1))
)

nrow(detections)
```

## Before fitting: what does this data support?

`detection_structure()` reads the table and reports which analyses it admits.
Nothing is fitted.

```{r}
detection_structure(detections)
```

Two lines there matter more than the rest.

**The nearest detection is not at zero.** That is the blind spot, and it decides
the model set below.

**Perception bias is not estimable.** This survey has one observer team, so no
amount of data recovers it. That is a property of the survey programme rather
than of any archive the data ends up in, and it does not improve with sample
size. The point of asking now is that fitting anyway produces a number, and the
number is wrong by a factor.

## The model set

An aircraft that cannot see beneath itself has two honest treatments, and they
are alternatives rather than a pair:

- a **gamma** key, which is unimodal with its peak away from zero, or
- **left truncation**, which removes the blind spot from the data.

Using both counts the blind spot twice, and `sweep_models()` warns when a model
set containing gamma is fitted with `left`.

```{r}
models <- model_set(key = c("hn", "hr", "gamma"))
models[, c("model_id", "key", "formula")]
```

## The sweep

Truncation is one value for the whole sweep, by construction. AIC compares
models fitted to the same data, so a table mixing truncations is not a ranking.

```{r, eval = has_mrds}
sw <- sweep_models(detections, models, truncation = 2500)
```

```{r, eval = has_mrds}
sw
```

## Read the table across, not down

The AIC column answers "which model best describes these distances". It does not
answer "what does this imply about how many animals there were", and the two can
disagree.

```{r, eval = has_mrds}
selection_table(sw)[, c("model_id", "delta_aic", "p", "p_cv", "esw", "cvm_p")]
```

`esw` is the effective strip half-width — the width of the strip that would have
produced this many detections under certain detection. It is what abundance is
divided by, so a spread in `esw` is a spread in density, directly and
proportionally. Across these three it runs from roughly 640 to 890 m, so the
same survey implies densities differing by about 40% depending on which model
you take.

Now read `cvm_p`, whose null is that the fitted function describes the observed
distances. **Every model in this table fails it**, the winner included. The AIC
column ranked them against each other and had no way to say that none of them
fit.

For binned data the column to read is `chisq_p` instead, because Cramér-von
Mises tests an empirical distribution that binned distances do not have.

### What the failure is telling you

It is not that a fourth key would fit better. It is that a **hard geometric
cutoff is not a detection function**. The gamma key models detection *falling
off* toward the trackline, smoothly; the aircraft here sees nothing at all below
100 m. No key function reproduces a cliff, so all of them misfit the near
distances and the goodness-of-fit test says so.

The fix is not a better model. It is to stop asking the model to describe
geometry.

## Left truncation, the other route

Remove the blind spot from the data rather than modelling it. Gamma is left out
of this set deliberately: it and `left` are alternative treatments of the same
phenomenon, and applying both counts the blind spot twice — `sweep_models()`
warns if you try.

```{r, eval = has_mrds}
sw_left <- sweep_models(detections, model_set(c("hn", "hr")),
                        truncation = 2500, left = 100, quiet = TRUE)
selection_table(sw_left)[, c("model_id", "delta_aic", "p", "esw", "cvm_p")]
```

The half-normal now passes comfortably. Same survey, same animals, and the
difference between a model set that fits and one that does not was a statement
about the aircraft rather than about the whales.

Note that `hr` still fails, so the goodness-of-fit column is doing work here
beyond rubber-stamping the winner.

Comparing these AICs with the ones in the previous table would be a mistake —
different data, different likelihood. What *is* comparable is what the two
routes imply for `esw`, and that is a question about the survey rather than
about the models.

## Covariates

A covariate model fits a separate detection function per covariate value, so
`p̄` becomes a mean over the animals seen.

```{r, eval = has_mrds}
sw_cov <- sweep_models(
  detections,
  model_set(key = "hn", formula = list(~1, ~beaufort)),
  truncation = 2500, left = 100, quiet = TRUE
)
selection_table(sw_cov)[, c("model_id", "delta_aic", "p", "p_cv", "esw")]
```

Sea state was built into the simulation, so a model that knows about it should
win here. On real data it often does not, and a covariate that does not earn its
parameter is worth dropping.

## Plotting

`effect_estimates.ddf()` makes a fitted detection function plottable with
[fancyfx](https://github.com/chross22/fancyfx), giving the standard diagnostic —
fitted g(x) with a delta-method ribbon and a rug of the observed distances.

```{r, eval = FALSE}
fancyfx::plotEffects(sw$fits[["hn"]], dat = detections, var = "distance")
```

For a **covariate** model, the area under that curve is not the `esw` in the
table. `mrds` reports `average.p` as a Horvitz-Thompson mean, weighting each
animal by the inverse of its own detection probability; the plotted curve is the
plain mean over the animals as observed. They coincide only when detection
probability is constant.

## g(0): what the fit assumed

Every fit above conditions on the animal having been available and seen. A
mis-specified `g(0)` scales every candidate in the set by the same factor, so
the ranking looks untouched while the density is wrong — which is exactly why it
cannot be caught by model selection and has to be handled deliberately.

Three separate things present as `g(0) < 1`:

| component | what it is | where it comes from |
|---|---|---|
| blind spot | the aircraft cannot see beneath itself | handled above, by gamma or `left` |
| availability | the animal was submerged | dive data — external |
| perception | the animal was up and missed | a second observer team |

### Availability is computed, not estimated

An animal submerged for the whole pass is missed at *every* distance equally, so
availability leaves no signature in the distance distribution. There is nothing
to estimate it from. It is computed instead, from dive intervals and how long
the platform keeps a point in view.

```{r}
ganley_availability
```

Those are measurements from Cape Cod Bay, and the swing is the point: **0.27 in
January against 0.91 in April**, as the copepods the whales feed on move up
through the water column. A single representative figure would be a real number
for one month and wrong by a factor of three for another.

`availability()` computes it from your own platform and dive data. An aircraft's
field of view is a wedge running forward and aft, so time in view *grows* with
perpendicular distance — `view_window_aerial()` is that geometry, and
`view_window()` is the circular case, which narrows.

```{r}
avail <- availability(
  surface    = c(48, 226, 67, 801),   # mean surfacing interval, seconds
  dive       = c(533, 256, 219, 88),  # mean diving interval, seconds
  window     = view_window_aerial(trackline = 51.22, speed = 51.4, slope = 0.03),
  se_surface = c(6, 20, 8, 60),
  se_dive    = c(40, 25, 20, 9),
  key        = c("Jan", "Feb", "Mar", "Apr")
)
avail
```

### Assembling the correction

`g0()` stacks the components, multiplies them and propagates the variance. It
builds an object and hands it off; this package does not compute abundance, so
it does not apply it.

```{r}
avail$source <- "focal follows, this survey"

perception <- data.frame(
  key = NA_character_, component = "perception", value = 0.68, se = 0.09,
  source = "borrowed from a double-observer programme - not measured here"
)

g0(avail, perception)
```

Three things that object refuses to let you do quietly:

**It will not default to 1.** Leave perception out and it says so, in a warning
and again every time it prints:

```{r}
g0(avail)
```

**It will not accept a component without a standard error.** The CV of `g(0)`
routinely dominates the CV of abundance, so dropping it inverts which
uncertainty matters. `availability()` returns `se = NA` when you give it no
interval standard errors, and `g0()` refuses that — the two rules close a loop.

**It will not join components keyed on different things.** Availability by month
and perception by year is the common case, and it has no correct join: the
answer is a value per month-year. Build that cross product deliberately.

Under independence the squared CVs add, `CV(g0)² = Σ CV(xᵢ)²`, so the least
precise component sets the floor. A perception estimate with a 30% CV cannot be
rescued by an availability estimate with a 2% one.

## Where this stops

Divide abundance by `g(0)` and carry its CV through — at the abundance step, in
the analysis layer, not here. `dsfit` fits detection functions and reports what
they imply; density surface models and their covariate stack live in their own
package.

And it does not estimate `g(0)`. Correcting for one component while believing
you have corrected for another is how these estimates go wrong by a factor
rather than a percentage, which is why the components are named separately all
the way through.
